---
title: "Demonsrate Usage of `f_vhm_agg()`"
author: "Attilio Benini, attilio.benini@bfh.ch"
date: "2024-09-20"
# date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: show
    toc: true
    number_sections: no
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, fig.align='left')
Sys.setenv(LANG = "en_US.UTF-8")
```

## Introduction

_**TBk**_ issue [#45](https://github.com/HAFL-WWI/TBk/issues/45) raises the
question of an alternative to the current method of (pre-)processing from input
VHM to VHM with resolution (10m X 10m). The issue mentions an earlier alternative
designed for the Canton of Glarus and written in _**R**_ 
([vhm_resample.R](https://github.com/HAFL-WWI/TBk/files/15335154/vhm_resample.R.txt)).

That script contained a function `vhm3()`, which served as specification of
[`raster::aggregate()`](https://search.r-project.org/CRAN/refmans/raster/html/aggregate.html)'s 
argument `fun`. Building on `vhm3()` I've written `vhm_agg()`. At the time 
`vhm_agg()`'s functionality hardly differs from one of its precursor. The main 
differences are:

* tidied up code improving readability.
* avoidance of redundancies.
* all parameters relevant for defining cases 1 to 5 within the implemented 
decision tree are included as function arguments.

This facilitate experimenting and further development.

But overall `vhm_agg()` in its current version is actually too complex. The built
in 5-step decision tree includes two type of criteria:

1) metrics derived from values of the original / detailed VHM within each 
10m x 10m aggregation pixel.
2) absolute values assigning VHM values to different classes of height / canopy
layers.

This complexity makes it hard to understand and explain `vhm_agg()`. This goes for
developers and users. Thus it's also doubtful, if this could ever be translated into
a user-friendly GUI, which should be _**TBk**_'s aim. Let alone the difficulty of
implementing a raster aggregation with a build in decision into pyqgis.

Anyhow, before going to QGIS / pyqgis research on a reasonable alternative of VHM
aggregation to 10m x 10m pixels needs to done. Apart form some very few meters
above ground, we should avoid using absolute values as thresholds. I see potential
for statistic driven metrics like quantiles and the [Gini coefficient](https://en.wikipedia.org/wiki/Gini_coefficient).
For the later there's also an _**R**_ function: [`DescTools::Gini()`](https://andrisignorell.github.io/DescTools/reference/Gini.html).

Below I demonstrate how to use `vhm_agg()` with [`terra::aggregate()`](https://search.r-project.org/CRAN/refmans/terra/html/aggregate.html)'.
It probably also works with `raster` (not tested yet). However, I don't see any
reason for still using [outdated _**R**_-packages](https://geocompx.org/post/2023/rgdal-retirement/)
like `raster`.

If you want to try out `vhm_agg()`, have a look at the description of the function's
arguments included in the script `vhm_agg.R`.

In order to get a proper result when using `vhm_agg()` in conjunction with 
`terra::aggregate()` some points have to be considered:

1)  `terra::aggregate()`'s argument `fact` (= number of cells in each direction (x / y)
which are aggregated into the output-pixels) must be specified. Since the aggregated
VHM must have a resolution of 10m X 10m, the original VHM must have an according
resolution. This is given when dividing 10m by the resolution of original VHM
results in remainder of zero. If this precondition is not met, the original VHM
must resampled to an intermediate with resolution fit for aggregating to the 
aspired 10m X 10m output.

2) Aligning the aggregated VHM (10m X 10m) to the origin (X, Y) = (0, 0),
is [_**TBk**_'s new preprocessing default](https://github.com/HAFL-WWI/TBk/pull/57).
Aligning to the origin (0, 0) means all pixel edges match coordinates = k * pixel-resolution,
where k is an integer. If this is not true for the original VHM, it must be shifted
to fulfill also this precondition. Furthermore the extent of the VHM used as input
for `terra::aggregate()` must fit to 10m x 10m pixels aligned with origin (0, 0).

3)  Finally the input VHM must be masked with the same forest mask used else in
the _**TBk**_-project.

4) The input VHM should be cleaned, meaning not including values < 0 and > max. 
possible tree height.

Below, the as input used VHM does matches most the a.m. criteria resp. can be 
processed with little effort to match them. Some tricks to check and achieve point
1) and 2) are presented. If you need help or plan using `vhm_agg()` / 
`terra::aggregate()` with larger input volumes, do contact me.

## Start time 

```{r}
t_start <- Sys.time()
```

## R-packages

R-packages from CRAN needed:

```{r, message=FALSE, warning=FALSE}
# library(tidyverse)
library(sf)
library(terra)
library(mapview) 
# library(tmap)
```

R-external dependencies versions of the libraries linked to `sf`-package:

```{r}
sf_extSoftVersion()
```

## Paths

Relative path to script with function `f_vhm_agg()`:

```{r}
path_f_vhm_agg <- file.path(getwd(), "f_vhm_agg.R")
file.exists(path_f_vhm_agg)
```

Note that, here we assume that this script (`f_vhm_agg_demo.Rmd`) and `f_vhm_agg.R`
are both located in the same (working) directory.

Absolute paths for input data:

```{r}
path_forest_mask <- file.path("P:", "HAFL", "7 WWI", "78 Temp", "780 Temp_Mittelbau", "Temp_Benini", "test_f_vhm_agg", "waldmaske_hafl.gpkg")
file.exists(path_forest_mask)

path_vhm <- file.path("P:", "HAFL", "7 WWI", "78 Temp", "780 Temp_Mittelbau", "Temp_Benini", "test_f_vhm_agg", "VHM_clean_hafl.tif")
file.exists(path_vhm)
```

## Load `f_vhm_agg()` 

```{r}
source(path_f_vhm_agg)
```

## Helfper-`function()`

Function for mapping bounding box of a spatial object in a interactive map
([`mapview`](https://r-spatial.github.io/mapview/)):

```{r}
f_map_bb <- function(x, layer.name, col = "red", lwd = 1.5, native.crs = FALSE){
  st_bbox(x) %>% 
    st_as_sfc() %>% 
    st_sf() %>%
    mapview(
      lwd           = lwd,
      color         = col,
      col.region    = col,
      alpha.regions = 0,
      layer.name    = layer.name,
      native.crs    = native.crs
      )
}
```

## Get data

Forest mask:

```{r}
forest_mask <- terra::vect(path_forest_mask)
```

High-resolution VHM input :

```{r}
vhm <- terra::rast(path_vhm)
```

View inputs:

```{r out.width="60%", fig.cap="High-resolution input VHM and forest mask."}
plot(vhm)
plot(forest_mask, border = "red", lwd = 2, add = TRUE)
```

$\Longrightarrow$ The input VHM is yet not not masked, which is good in case we
have to apply shifting and/or resampling to it!

## Extent of VHM 10m 

The extent / bounding box of the aggregated VHM 10m must

* cover the whole forest mask and
* when divided by 10m all remainders must be zero

```{r}
# bounding box of forest mask
bb_forest_mask <- st_bbox(forest_mask)

# bounding box of VHM 10m, must include whole forest mask and ...
bb_vhm_10m <-
  st_bbox(
    c(
      floor(bb_forest_mask$xmin / 10) * 10,
      floor(bb_forest_mask$ymin / 10) * 10,
      ceiling(bb_forest_mask$xmax / 10) * 10,
      ceiling(bb_forest_mask$ymax / 10) * 10
    ),
    crs = st_crs(bb_forest_mask)
  )

bb_vhm_10m - bb_forest_mask

# when divided by 10m all remainders must be zero
all(bb_vhm_10m %% 10 == 0)
# --> desired alignment of VHM 10 output
```

## Check VHM input

```{r}
# resolution of of original VHM 
res(vhm)

# dividing the resolution of VHM 10m by resolution of original VHM ... 
10 %% unique(res(vhm)) == 0
# ... must result only in remainder = 0

# bounding box of original VHM
bb_vhm <- st_bbox(vhm)

# dividing the bounding box of the original VHM by its resolution ...
bb_vhm %% unique(res(vhm)) == 0
# ... must result only in remainders = 0
```

## Prepare VHM as aggregation input

Since the above checks have revealed that the input VHM needs no resampling and/or 
shifting, we can mask it with the forest mask:

```{r out.width="60%", fig.cap="Masked high-resolution VHM."}
vhm <- terra::mask(vhm, forest_mask)
plot(vhm)
```

How does the bounding boxes of the original VHM overlap with that of the VHM 10m?

```{r out.width="60%", fig.cap="Forest mask and boundigng boxes of orignal VHM and VHM 10m."}
bb_vhm_10m - bb_vhm  

mapview(forest_mask, col.region = "forestgreen") +
  # f_map_bb(forest_mask, "forest mask", "forestgreen") +
  f_map_bb(bb_vhm, "bb original VHM", "blue") + 
  f_map_bb(bb_vhm_10m, "bb VHM 10m")
```

If the bounding box of the original VHM goes on any side outside of extent of
the VHM 10m, we crop the original VHM with the bounding box / extent of VHM 10m.

```{r}
bb_vhm_diff <- bb_vhm_10m - bb_vhm
outside_bb_vhm_10m <- any(bb_vhm_diff[c("xmin", "ymin")] > 0, bb_vhm_diff[c("xmax", "ymax")] < 0)
outside_bb_vhm_10m
if(outside_bb_vhm_10m){
  vhm <- crop(vhm, ext(bb_vhm_10m))
}
```

If (after cropping) the extent of VHM 10m goes beyond the extent of the original
VHM: 

```{r}
any(bb_vhm_10m != st_bbox(vhm))

if(any(bb_vhm_10m != st_bbox(vhm))) { 
  # 1) get the extending distances on all 4 sides (left, right, bottom, top) ...
  extent_dist <- unname(abs(bb_vhm_10m - st_bbox(vhm))[c("xmin", "xmax", "ymin", "ymax")])
  # 2) get number of cells to be added on each side ...
  extent_n <- extent_dist / res(vhm)[1]
  # 3) finally extend original / detailed VHM to extent of VHM 10m  
  vhm <- extend(vhm, extent_n)
}
```

Check if extent of high-resolution VHM and VHM 10m are now identical:

```{r}
bb_vhm_10m == st_bbox(vhm)
bb_vhm_10m - st_bbox(vhm)
```

## Aggreate VHM 10m

Get aggregation factor by dividing 10m (the resolution of aggregated VHM) with
resolution of of high-resolution VHM:

```{r out.width="60%", fig.cap="Masked high-resolution VHM."}
# resolution of of high-resolution VHM
terra::res(vhm)
# get aggregation factor
(fact <- 10 / terra::res(vhm)[1]) 
```

Check if aggregation factor is an integer:

```{r}
fact %% 1 == 0
```

Finally aggregate with `f_vhm_agg()`:

```{r out.width="60%", fig.cap="Aggregaed VHM 10m."}
vhm_10m <- terra::aggregate(vhm, fact = fact, fun = f_vhm_agg)
names(vhm_10m) <- "VHM_agg_10m"
plot(vhm_10m)
```

Note that above `f_vhm_agg()` is used with default argument settings. Non-default
argument specifications can be passed on to `fun` (`= f_vhm_agg`) like this:

```{r eval=FALSE}
terra::aggregate(
  vhm,
  fact             = fact,
  fun              = f_vhm_agg,
  f_h_top_min_1    = function(h_max){max(0.85 * h_max - 3, 1)}, # max(,1) ensures it is > 0
  min_percentage_1 = 40,
  f_h_top_min_2    = function(h_max){max(0.8 * h_max - 3, 1)}, # max(,1) ensures it is > 0
  min_percentage_2 = 55
  )
```

Reminder: Have a look at the description of the `f_vhm_agg()`'s arguments
included in the script `vhm_agg.R`.
